---
title: "Chapter 3 Problem 3.6a"
author: "Lizhuo ZHOU 20307100132"
output: pdf_document
date: '2023-10-22'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Problem 3.6a
# n 		= number of observations in the data set
# m 		= number of predictors in the data set
# chromosome = matrix of P individuals for a generation where each row
# 		  is a chromosome included for the model
# 		  (1 = father, 0 = mother)
# chromosome.next 	= matrix of P individuals for the new generation where each
# 		  row is a chromosome included for the model
# 		  (1 = father, 0 = mother)
# ordering.PML 	= PML values for the models of each generation
# P 		= size of each generation
# itr 		= number of generations to run
# m.rate 	= mutation rate
# r 		= ranks of PMLs for a generation
# phi 		= fitness values for a generation
# ordering = vector of a chromosome included in the best model found
# 		  (1 = father, 0 = mother)
# best.PML 	= PML value for the best model found
# PML 		= PML values for the models at each step

```


```{r}
#First read the data, set the number of generations and the mutation rate
rm(list=ls()) 
genedata<-read.table("geneticmapping.dat",header=TRUE)
P<-96
m.rate<-0.01
n<-ncol(genedata)
parent<-matrix(0,nrow=P,ncol=n)
m = length(genedata[1,])
itr = 10000
r = matrix(0,P,1)
phi = matrix(0,P,1)
chromosome = matrix(0,P,m)
chromosome.next = matrix(0,P,m)
ordering.PML = matrix(0,P,1)
PML = matrix(0,P,itr)
ordering = NULL
best.PML = -100000
best.PML.gen = rep(0,itr)

#Define a function that calculates genetic distances
d<-function(x,i,j){
  ii<-x[i]
  jj<-x[j]
  dis<-abs(as.numeric(ii-jj))/96
  return(dis)
}


#Define a function that calculates the maximal profile likelihood
Profile_Likelihood<-function(i){
  s<-0
  for(j in 1:11){
    if(d(i,j,j+1)==1){
      s<-s
    }else if(d(i,j,j+1)==0){
      s<-s
    }else{
      s<- s+96*((1-d(i,j,j+1))*log(1-d(i,j,j+1))+d(i,j,j+1)*log(d(i,j,j+1)))
      }
  }
  return(s)
}

# Function to perform order crossover
order_crossover <- function(parent1, parent2,pos) {
  # Choose two random crossover points
  child_1 <- rep(NA, 12)
  #child_2 <- rep(NA, 12)
  if (pos==1){
    child_1[2:length(parent1)] <- parent1[2:length(parent1)]
    child_1[1]<-parent2[1]
    #child_2[2:length(parent2)] <- parent2[2:length(parent2)]
    #child_2[1]<-parent1[1]
  }else{
    child_1[1:pos]<-parent1[1:pos]
    child_1[pos+1:length(parent1)]<-parent2[pos+1:length(parent2)]
    #child_2[1:pos]<-parent2[1:pos]
    #child_2[pos+1:length(parent2)]<-parent1[pos+1:length(parent1)]
  }
  child_1<-child_1[1:12]
  #child_2<-child_2[1:12]

  return(child_1)
}

#Experiment 1

# INITIALIZES STARTING GENERATION, FITNESS VALUES
set.seed(321321123)
genedata<-matrix(unlist(genedata),nrow=96,ncol=12)
first_GEN<-genedata[sample(1:nrow(genedata),P),]
for(i in 1:P){
	chromosome[i,]<-first_GEN[i,]
	ordering.PML[i] <-Profile_Likelihood(chromosome[i,])
}
if(sum(ordering.PML) >best.PML){
			ordering <- chromosome
			best.PML <- sum(ordering.PML)
		}
r = rank(ordering.PML)
phi = 2*r/(P*(P+1))
best.PML.gen[1]=best.PML


## MAIN
for(j in 1:itr-1){

	# BUILDS THE NEW GENERATION, SELECTING FIRST PARENT BASED ON
	# FITNESS AND THE SECOND PARENT AT RANDOM
	for(i in 1:96){
		parent.1 = chromosome[sample(1:P,1,prob=phi),]
		parent.2 = chromosome[sample(1:P,1),]
		#parent.2 = chromosome[-((chromosome[sample(1:P,1),])%in% chromosome),][sample(1:P-1,1)]
		pos = sample(1:(m-1),1)
		mutate = rbinom(m,1,m.rate)
		chromosome.next[i,] = order_crossover(parent.1,parent.2,pos)
		chromosome.next[i,] = (chromosome.next[i,]+mutate)%%2
	}
	chromosome = chromosome.next

	# UPDATES PML VALUES, FITNESS VALUES FOR NEW GENERATION
	for(i in 1:P){
		ordering.PML[i] = Profile_Likelihood(chromosome[i,])
		PML[i,j+1] = ordering.PML[i]
		if(ordering.PML[i] >best.PML){
			ordering = chromosome[i,]
			best.PML = ordering.PML[i]
		}
	}
	
	if(sum(ordering.PML) >best.PML){
			ordering = chromosome
			best.PML = sum(ordering.PML)
		}
	best.PML.gen[j+1]=best.PML
	r = rank(ordering.PML)
	phi = 2*r/(P*(P+1))
}

## OUTPUT
chromosome 		# BEST LIST OF PREDICTORS FOUND
best.PML 	# PML VALUE

```
```{r}
#Experiment 2

# INITIALIZES STARTING GENERATION, FITNESS VALUES
set.seed(321328342)
genedata<-matrix(unlist(genedata),nrow=96,ncol=12)
first_GEN<-genedata[sample(1:nrow(genedata),P),]
for(i in 1:P){
	chromosome[i,]<-first_GEN[i,]
	ordering.PML[i] <-Profile_Likelihood(chromosome[i,])
}
if(sum(ordering.PML) >best.PML){
			ordering <- chromosome
			best.PML <- sum(ordering.PML)
		}
r = rank(ordering.PML)
phi = 2*r/(P*(P+1))
best.PML.gen[1]=best.PML


## MAIN
for(j in 1:itr-1){

	# BUILDS THE NEW GENERATION, SELECTING FIRST PARENT BASED ON
	# FITNESS AND THE SECOND PARENT AT RANDOM
	for(i in 1:96){
		parent.1 = chromosome[sample(1:P,1,prob=phi),]
		parent.2 = chromosome[sample(1:P,1),]
		#parent.2 = chromosome[-((chromosome[sample(1:P,1),])%in% chromosome),][sample(1:P-1,1)]
		pos = sample(1:(m-1),1)
		mutate = rbinom(m,1,m.rate)
		chromosome.next[i,] = order_crossover(parent.1,parent.2,pos)
		chromosome.next[i,] = (chromosome.next[i,]+mutate)%%2
	}
	chromosome = chromosome.next

	# UPDATES PML VALUES, FITNESS VALUES FOR NEW GENERATION
	for(i in 1:P){
		ordering.PML[i] = Profile_Likelihood(chromosome[i,])
		PML[i,j+1] = ordering.PML[i]
		if(ordering.PML[i] >best.PML){
			ordering = chromosome[i,]
			best.PML = ordering.PML[i]
		}
	}
	
	if(sum(ordering.PML) >best.PML){
			ordering = chromosome
			best.PML = sum(ordering.PML)
		}
	best.PML.gen[j+1]=best.PML
	r = rank(ordering.PML)
	phi = 2*r/(P*(P+1))
}

## OUTPUT
chromosome 		# BEST LIST OF PREDICTORS FOUND
best.PML 	# PML VALUE



```

```{r}
#Experiment 3

# INITIALIZES STARTING GENERATION, FITNESS VALUES
set.seed(3241623)
genedata<-matrix(unlist(genedata),nrow=96,ncol=12)
first_GEN<-genedata[sample(1:nrow(genedata),P),]
for(i in 1:P){
	chromosome[i,]<-first_GEN[i,]
	ordering.PML[i] <-Profile_Likelihood(chromosome[i,])
}
if(sum(ordering.PML) >best.PML){
			ordering <- chromosome
			best.PML <- sum(ordering.PML)
		}
r = rank(ordering.PML)
phi = 2*r/(P*(P+1))
best.PML.gen[1]=best.PML


## MAIN
for(j in 1:itr-1){

	# BUILDS THE NEW GENERATION, SELECTING FIRST PARENT BASED ON
	# FITNESS AND THE SECOND PARENT AT RANDOM
	for(i in 1:96){
		parent.1 = chromosome[sample(1:P,1,prob=phi),]
		parent.2 = chromosome[sample(1:P,1),]
		#parent.2 = chromosome[-((chromosome[sample(1:P,1),])%in% chromosome),][sample(1:P-1,1)]
		pos = sample(1:(m-1),1)
		mutate = rbinom(m,1,m.rate)
		chromosome.next[i,] = order_crossover(parent.1,parent.2,pos)
		chromosome.next[i,] = (chromosome.next[i,]+mutate)%%2
	}
	chromosome = chromosome.next

	# UPDATES PML VALUES, FITNESS VALUES FOR NEW GENERATION
	for(i in 1:P){
		ordering.PML[i] = Profile_Likelihood(chromosome[i,])
		PML[i,j+1] = ordering.PML[i]
		if(ordering.PML[i] >best.PML){
			ordering = chromosome[i,]
			best.PML = ordering.PML[i]
		}
	}
	
	if(sum(ordering.PML) >best.PML){
			ordering = chromosome
			best.PML = sum(ordering.PML)
		}
	best.PML.gen[j+1]=best.PML
	r = rank(ordering.PML)
	phi = 2*r/(P*(P+1))
}

## OUTPUT
chromosome 		# BEST LIST OF PREDICTORS FOUND
best.PML 	# PML VALUE



```
```{r}
#Experiment 4

# INITIALIZES STARTING GENERATION, FITNESS VALUES
set.seed(32000623)
genedata<-matrix(unlist(genedata),nrow=96,ncol=12)
first_GEN<-genedata[sample(1:nrow(genedata),P),]
for(i in 1:P){
	chromosome[i,]<-first_GEN[i,]
	ordering.PML[i] <-Profile_Likelihood(chromosome[i,])
}
if(sum(ordering.PML) >best.PML){
			ordering <- chromosome
			best.PML <- sum(ordering.PML)
		}
r = rank(ordering.PML)
phi = 2*r/(P*(P+1))
best.PML.gen[1]=best.PML


## MAIN
for(j in 1:itr-1){

	# BUILDS THE NEW GENERATION, SELECTING FIRST PARENT BASED ON
	# FITNESS AND THE SECOND PARENT AT RANDOM
	for(i in 1:96){
		parent.1 = chromosome[sample(1:P,1,prob=phi),]
		parent.2 = chromosome[sample(1:P,1),]
		#parent.2 = chromosome[-((chromosome[sample(1:P,1),])%in% chromosome),][sample(1:P-1,1)]
		pos = sample(1:(m-1),1)
		mutate = rbinom(m,1,m.rate)
		chromosome.next[i,] = order_crossover(parent.1,parent.2,pos)
		chromosome.next[i,] = (chromosome.next[i,]+mutate)%%2
	}
	chromosome = chromosome.next

	# UPDATES PML VALUES, FITNESS VALUES FOR NEW GENERATION
	for(i in 1:P){
		ordering.PML[i] = Profile_Likelihood(chromosome[i,])
		PML[i,j+1] = ordering.PML[i]
		if(ordering.PML[i] >best.PML){
			ordering = chromosome[i,]
			best.PML = ordering.PML[i]
		}
	}
	
	if(sum(ordering.PML) >best.PML){
			ordering = chromosome
			best.PML = sum(ordering.PML)
		}
	best.PML.gen[j+1]=best.PML
	r = rank(ordering.PML)
	phi = 2*r/(P*(P+1))
}

## OUTPUT
chromosome 		# BEST LIST OF PREDICTORS FOUND
best.PML 	# PML VALUE




```

```{r}
#Experiment 5

# INITIALIZES STARTING GENERATION, FITNESS VALUES
set.seed(12382766)
genedata<-matrix(unlist(genedata),nrow=96,ncol=12)
first_GEN<-genedata[sample(1:nrow(genedata),P),]
for(i in 1:P){
	chromosome[i,]<-first_GEN[i,]
	ordering.PML[i] <-Profile_Likelihood(chromosome[i,])
}
if(sum(ordering.PML) >best.PML){
			ordering <- chromosome
			best.PML <- sum(ordering.PML)
		}
r = rank(ordering.PML)
phi = 2*r/(P*(P+1))
best.PML.gen[1]=best.PML


## MAIN
for(j in 1:itr-1){

	# BUILDS THE NEW GENERATION, SELECTING FIRST PARENT BASED ON
	# FITNESS AND THE SECOND PARENT AT RANDOM
	for(i in 1:96){
		parent.1 = chromosome[sample(1:P,1,prob=phi),]
		parent.2 = chromosome[sample(1:P,1),]
		#parent.2 = chromosome[-((chromosome[sample(1:P,1),])%in% chromosome),][sample(1:P-1,1)]
		pos = sample(1:(m-1),1)
		mutate = rbinom(m,1,m.rate)
		chromosome.next[i,] = order_crossover(parent.1,parent.2,pos)
		chromosome.next[i,] = (chromosome.next[i,]+mutate)%%2
	}
	chromosome = chromosome.next

	# UPDATES PML VALUES, FITNESS VALUES FOR NEW GENERATION
	for(i in 1:P){
		ordering.PML[i] = Profile_Likelihood(chromosome[i,])
		PML[i,j+1] = ordering.PML[i]
		if(ordering.PML[i] >best.PML){
			ordering = chromosome[i,]
			best.PML = ordering.PML[i]
		}
	}
	
	if(sum(ordering.PML) >best.PML){
			ordering = chromosome
			best.PML = sum(ordering.PML)
		}
	best.PML.gen[j+1]=best.PML
	r = rank(ordering.PML)
	phi = 2*r/(P*(P+1))
}

## OUTPUT
chromosome 		# BEST LIST OF PREDICTORS FOUND
best.PML 	# PML VALUE




```
\
\ \ Conclusion:
\
\ \ From Experiment 1-5 above, we could see that the results produced by genetic algorithm after multiple iterations are dominated by 'strong' parent generations(in this case, parent generations with large Profile Likelihood).
\
\ \ Performance of the Algorithm:
\
\ \ Robust, while lack of diversity.
\
\ \ Ideas for improved reseach:
\
\ \ Try different mutation rates, times of iterations and size of a generations.

