---
title: "Chapter 9_Problem 9.7"
author: "Lizhuo ZHOU 20307100132"
date: "2023-12-18"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Part I: Use the problem of estimating the mean of a standard Cauchy distribution to illustrate how the bootstrap can fail for heavy-tailed distributions

\
\ \ 

```{r}
# Part I

set.seed(122)
n <- 1000
original_sample <- rcauchy(n)
num_bootstraps <- 1000
bootstrap_means <- numeric(num_bootstraps)

# Perform the bootstrap resampling
for (i in 1:num_bootstraps) {
  bootstrap_sample <- sample(original_sample, size = n, replace = TRUE)
  bootstrap_means[i] <- mean(bootstrap_sample)
}

# Plot the distribution of bootstrap sample means
hist(bootstrap_means, breaks = 50, main = "Bootstrap Means of Standard Cauchy Samples", xlab = "Mean")

cat("Bootstrap Mean of Means:", mean(bootstrap_means), "\n")
cat("Bootstrap SD of Means:", sd(bootstrap_means), "\n")

```

## Part II: Use the problem of estimating $\theta$ for the Unif(0,$\theta$) distribution to illustrate how the bootstrap can fail for extremes.

```{r}
# Part II
set.seed(1774)

# True parameter theta
theta <- 15

# Generate a sample of size n from a Unif(0, theta) distribution
n <- 100
original_sample <- runif(n, 0, theta)

num_bootstraps <- 1000

# Initialize a vector to store the bootstrap estimates of theta (maximum values)
bootstrap_maxima <- numeric(num_bootstraps)

# Perform the bootstrap resampling
for (i in 1:num_bootstraps) {
  bootstrap_sample <- sample(original_sample, size = n, replace = TRUE)
  bootstrap_maxima[i] <- max(bootstrap_sample)
}

# Plot the distribution of bootstrap estimates for theta
hist(bootstrap_maxima, breaks = 50, main = "Bootstrap Estimates of Theta", xlab = "Estimated Theta")

# Calculate and print the mean of the bootstrap estimates of theta
bootstrap_theta_mean <- mean(bootstrap_maxima)
cat("Bootstrap Mean of Theta Estimates:", bootstrap_theta_mean, "\n")

# Calculate and print the bias and the standard deviation of the bootstrap estimates
bootstrap_bias <- bootstrap_theta_mean - theta
bootstrap_sd <- sd(bootstrap_maxima)
cat("Bootstrap Bias of Theta Estimates:", bootstrap_bias, "\n")
cat("Bootstrap SD of Theta Estimates:", bootstrap_sd, "\n")


```
\
\ \ Conclusion:
\
\ \ For part I, with background knowledge we know that, a standard Cauchy distribution is an example of a heavy-tailed distribution, and it does not have a defined mean. From the example above, we could see the bootstrap method fail for heavy-tailed distributions, as this property indicates that extreme values are more common, thus these extreme values may heavily influence the sample mean, leading to high variability in the bootstrap estimates.
\
\ \ When applying the bootstrap to a sample from a Cauchy distribution and calculate the mean for each resample, you will likely get a very wide distribution of means that does not converge to any particular value, even as the sample size increases, as shown in the plot.
\
\ \ For part II, we could see that the bootstrap fails for extremes.When bootstrapping, each resample is taken from the original sample, meaning that the bootstrap samples will never contain a value larger than the maximum observed value M. As a result, the bootstrap distribution of $\theta$ will be biased downwards, as it cannot account for the possibility of larger values that could be observed in a new sample.

