---
title: "Chapter 9 Problem 1"
author: "Lizhuo ZHOU 20307100132"
date: "2023-12-18"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


1. Assume $\{X_{i}\},i=1,...,100$ is a sequence of iid standard normal random variables, and $Y_{i}$ satisfies the following regression$Y_{i}=X_{i}+\epsilon_{i},\epsilon_{i} \sim N(0,\sigma^{2})$. Suppose $\sigma^{2}=2$.

Please use paired bootstrap, bootstrap  residual and bootstrap error to compare the variance of regression coefficient estimates

\
\ \ From the statement, we could see that $Y_{i} \sim N(0,3),i=1,2,...,100$.
\
\ \ From the background knowledge of regression analysis, $\hat{\beta}=(X^{T}X)^{-1}X^{T}y$

```{r 1}
# First sample the independent variables
set.seed(10048)
X<-rnorm(100,0,1)
Y<-rnorm(100,0,3)
XY<-cbind(X,Y)
```

## Solution I: Paired Bootstrap

\
\ \ By definition of paired bootstrap, bootstrap $\{Z_{i}=(X_{i},Y_{i})\},i=1,2,...,n$

```{r 2}
# Solution I: Paired Bootstrap
# Sample size: 10 pairs each time; Sample times: 100

PB_coefficient_estimate<-c()
sample_PB_X<-matrix(0,nrow=10,ncol=1)
sample_PB_Y<-matrix(0,nrow=10,ncol=1)
beta_PB_estimate<-c()
sample_matrix<-matrix(0,nrow=100,ncol=10)

set.seed(100345)
for(i in 1:100){sample_matrix[i,]<-sample(1:100,10,replace=TRUE)}

for(j in 1:100){
  sample_index<-sample_matrix[j,]
  for (k in 1:10){
    sample_PB_X[k,1]<-XY[sample_index[k],1]
    sample_PB_Y[k,1]<-XY[sample_index[k],2]
  }
  beta_PB_estimate[j]<-(t(sample_PB_X)%*%sample_PB_X)^(-1)*t(sample_PB_X)%*%sample_PB_Y
  
  
}

# The variance of regression coefficient estimates——by paired bootstrap
var(beta_PB_estimate)


```


## Solution II: Bootstrap Residual
\
\ \ By definition of bootstrapping residual, $Y_{i,Boot}=Y_{i, Pred}+R_{rand}$
```{r 3}

# Solution II: Bootstrap Residual
# Sample size: 10 pairs each time; Sample times: 100
# Prepare samples
library(tidyverse)

df=tibble(
  ID=1:100,
  X=X,
  Y=Y)

df

```

```{r 4}
# Initial linear regression, save the estimated coefficient
fit_lm<-lm(Y~X,df)
b=coef(fit_lm)
b

```

```{r 5}
# Residual and estimated coefficient used for Bootstrap
res=tibble(
  ID=1:100,
  resid=fit_lm$residuals,
  pred=fit_lm$fitted.values)

res

```

```{r 6}
library(rsample)

set.seed(12345)

Residual_boostrap = bootstraps(res, times = 1000) %>% 
  mutate(data = map(splits, analysis)) %>% 
  select(-1) %>% 
  unnest(data) %>% 
  mutate(weightboot = resid + pred) %>% 
  left_join(df, by = "ID") 

Residual_boostrap
```

```{r 7}

rlt = Residual_boostrap %>% 
  group_nest(id) %>% 
  mutate(model = map(data, ~ lm(weightboot ~ X, .x)),
         coefs = map(model, coef)) %>% 
  unnest_wider(coefs)

rlt
```
```{r 8}

# The variance of regression coefficient estimates——by residual bootstrapping
var(rlt$X)

```

## Solution III: Bootstrap Error
\
\ \ By definition of bootstrapping error, first we estimate $\hat{\sigma^{2}}$, and then generate $\epsilon^{\star}_{i} \sim N(0,\hat{\sigma^{2}})$


```{r 9}
# Solution III: Bootstrap Error

# Estimated sigma_hat
sigma_hat<-var(res$resid)
# Generate a sequence of residual based on the estimated sigma_hat
set.seed(10057)
residual_star<-rnorm(100,0,sigma_hat^0.5)

error_bootstrap=tibble(
  ID=1:100,
  resid=residual_star,
  pred=fit_lm$fitted.values)

error_bootstrap


```

```{r 10}

set.seed(10089)

Error_boostrap = bootstraps(error_bootstrap, times = 1000) %>% 
  mutate(data = map(splits, analysis)) %>% 
  select(-1) %>% 
  unnest(data) %>% 
  mutate(bootstrapy = resid + pred) %>% 
  left_join(df, by = "ID") 

Error_boostrap

```


```{r 11}

rlt_error = Error_boostrap %>% 
  group_nest(id) %>% 
  mutate(model = map(data, ~ lm(bootstrapy ~ X, .x)),
         coefs = map(model, coef)) %>% 
  unnest_wider(coefs)

rlt_error


```



```{r 12}
# The variance of regression coefficient estimates——by bootstrapping error
var(rlt_error$X)

```
\
\ \ Conclusion:
\
\ \ Using paired bootstrap, the variance of the estimated regression coefficient is 1.149478;
\
\ \ Using bootstrap residual, the variance of the estimated regression coefficient is 0.08873295;
\
\ \ Using bootstrap error, the variance of the estimated regression coefficient is 0.0800907.
