---
title: "Chapter 6 Problem 6.8"
author: "Lizhuo ZHOU 20307100132"
date: "2023-11-09"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Consider the model given by $X \sim Lognormal(0,1)$ and $log Y= 9+3logX+\epsilon$, where $\epsilon \sim N(0,1)$. We wish to estimate $E\{Y/X\}$. Compare the performance of the standard Monte Carlo estimator and Rao-Blackwellized estimator.


```{r}
#Problem 6.8

num<-10000
set.seed(201)
X<-rlnorm(num,meanlog=0,sdlog=1)
epsilon<-rnorm(num,mean=0,sd=1)
Y<-exp(1)^(9+3*log(X)+epsilon)

h_x_SMA<-Y/X

# Standard Monte Carlo estimator

E_SMA<-sum(h_x_SMA)/length(h_x_SMA)


E_SMA

```
\
\ \ Next, perform the Rao-Blackwellization of the estimation $\hat{\mu_{RB}}=\dfrac{1}{n}\sum_{i=1}^{n}E\{h(X_{i1})|X_{i2}\}$.
\
\ \ In this scenario, we can obtain $X_{i}=(X_{i1},X_{i2})$, where $X_{i2}=log(X_{i1})$;
\
\ \ 
\begin{equation} \label{eq1}
\begin{split}
h(X_{i1}|X_{i2}) & = E\{Y/X|logX\} \\
 & = E\{exp\{log(Y/X)\}|logX\} \\
 & = E\{exp\{logY-logX\}|logX\} \\
 & = E\{exp\{9+2logX+\epsilon\}|logX\} \\
 & = exp(9.5+2logX)
\end{split}
\end{equation}
```{r}
# Rao-Blackwellized Estimator

X_i2<-log(X)

h_x_RB<-exp(9.5+2*X_i2)

E_RB<-sum(h_x_RB)/length(h_x_RB)

E_RB

```
\
\ \ Finally, let's solve the true value of $E\{Y/X\}$
\
\ \ 
\begin{equation} \label{eq2}
\begin{split}
E\{Y/X\} & = E\{exp\{log(Y/X)\} \\
 & = E\{exp\{logY-logX\}\} \\
 & = E\{exp\{9+2logX+\epsilon \}\} \\
 & = E\{exp\{11.5\}\} \\
 & = exp(11.5)
\end{split}
\end{equation}

```{r}
# True Value

exp(11.5)

```
\
\ \ Comparison:
\
\ \ From the above, we could see that Rao-Blackwellized estimator is closer to the true value compared with the standard Monte Carlo estimator.

